Poincaré and SimBio: a versatile and extensible Python ecosystem for modeling systems

### 2024-06-14 final/oup-authoring-template.tex
Move to final dir
Change cite to citep

Several tools already exist to \emph{define}, \emph{translate} and \emph{solve} \acp{CRN}.
BioSimulators.org \cite{shaikhBioSimulatorsCentralRegistry2022},
a registry of simulation tools,
lists at least 15 softwares categorized under Python including
COPASI \cite{hoopsCOPASICOmplexPAthway2006},
Tellurium \cite{choiTelluriumExtensiblePythonbased2018} and
PySB \cite{lopezProgrammingBiologicalModels2013}.
COPASI is a standalone software with a \ac{GUI}
that is widely used for its user-friendly interface and comprehensive features.
It also includes Python bindings (BASICO \cite{bergmannBASICOSimplifiedPython2023b}) that allow advanced scripting.
Tellurium is a Python-based modeling environment
that uses a C++ library called libRoadRunner \cite{welshLibRoadRunnerHighPerformance2023} in the backend to translate and solve models.
PySB is a Python library that created a \ac{DSL} using standard Python to define models,
which are then translated to \acp{ODE} using a Perl library called BioNetGen \cite{harrisBioNetGenAdvancesRulebased2016}.

One limitation of these tools is their extensibility from Python.
As they wrap libraries in other languages for defining, translating and/or solving models,
these steps cannot be altered or easily inspected from Python.
While they enable model definition and running simulations via Python scripts,
they cannot fully leverage Python's extensive package ecosystem.
For example,
COPASI and Tellurium do not allow the use of solvers defined in other Python packages,
and adding new integrators requires working with C++.
In particular, the step that translates into equations is not exposed by any of these tools.
As such, it is not possible to apply custom optimizations to the equations
or use automatic differentation packages such as JAX \cite{jax2018github} to compute the model's jacobian.

Another challenge is the way models are defined.
Many tools support the \ac{SBML} \cite{huckaSBMLL3V2} as an exchange format,
a \emph{de facto} standard for \acp{CRN} that defines species, parameters and reactions between species.
As writing \ac{SBML} directly is impractical,
Tellurium uses a \ac{DSL} called Antimony \cite{smithAntimonyModularModel2009} for defining models.
\Acp{DSL} allows to reuse the same code in different programming environments,
but are not recognized by default in \acp{IDE}
and, therefore, they cannot provide syntax highlighting, code completion, refactoring, and static analysis.
For Antimony, an extension providing these capabilities was developed for Visual Studio Code,
but its maintenance could be a demanding task for the systems biology community.
In the case of PySB,
using Python's dynamic nature,
it developers designed a \ac{DSL} within Python.
To save keystrokes,
it uses the global scope to create species and parameters,
without explicitly assigning them to Python variables or to the model,
but this approach is not fully compatible with \acp{IDE},
affecting the development experience.

@@

Modular code architecture makes code reusable, extensible, and easier to maintain.
Therefore, we split the code into three Python packages:
symbolite, to create symbolic expressions;
poincaré, to define dynamical systems;
and simbio, to define \acp{CRN} and interface with systems biology standards such as \ac{SBML}.
These are pure Python packages with standard dependencies from the PyData scientific stack
such as NumPy \cite{harrisArrayProgrammingNumPy2020} and pandas \cite{mckinneyDataStructuresStatistical2010}.
They are published in the \acf{PyPI},
where links to the source code and documentation hosted in GitHub can be found,
and can be easily installed with \texttt{pip\ install\ simbio},
which installs symbolite and poincare as dependencies.

Symbolite is a lightweight symbolics package to create algebraic mathematical expressions.
Unlike SymPy \cite{meurerSymPySymbolicComputing2017},
a widely-used Python library for symbolic mathematics,
it only provides the building of an expression tree
which can be inspected and compiled to various backends.
Symbolite is designed to facilitate the integration of new backends.
Currently, we have implementations for
NumPy \cite{harrisArrayProgrammingNumPy2020};
Numba \cite{lamNumbaLLVMbasedPython2015},
a \ac{JIT} compiler to LLVM;
SymPy \cite{meurerSymPySymbolicComputing2017};
and JAX \cite{jax2018github}, a
library that support automatic differentiation and compilation to \acp{GPU} and \acp{TPU}.

@@

For this last point,
\acp{IDE} that support \texttt{dataclass\_transform} \cite{debontePEP681Data2021}
can provide a tooltip with the expected signature (Figure~\ref{fig-ide}).
This requires the use of type annotations,
which play a more significant role in static type checking
as they can help to identify errors before running the code.
For instance, to parameterize the initial conditions of variables
we have to use a \texttt{Constant}.
If we try to use a \texttt{Parameter},
which could be a time-dependent expression,
it is flagged as a type error (Figure~\ref{fig-ide}).

@@

Several commonly used reactions are predefined as \texttt{MassAction} subclasses,
such as \texttt{MichaelisMenten} (\(S + E \leftrightarrow ES \rightarrow P + E\))
and its approximate form without the intermediate species \(ES\),
and it is also simple to implement used-defined ones
as subclasses of \texttt{RateLaw} or \texttt{MassAction}.
Additionally, SimBio supports importing models from \ac{SBML},
and downloading them directly from BioModels \cite{malik-sheriffBioModels15Years2020} (Figure~\ref{fig-simbio-io}).
Work is in progress to support exporting to \ac{SBML} and add more \ac{SBML} features.
Currently \ac{SBML} unsupported features include
algebraic rules, constraints, events, reactions with the fast attribute or with math stoichiometry,
units and compartments with size different from 1.

@@

To evaluate SimBio's reproducibility,
we used the \ac{SBML} test suite \cite{SBMLTestSuite},
which provides a set of \ac{SBML} models
and the expected result of a simulation.
Excluding models that use \ac{SBML} features not yet supported by SimBio,
every simulation returned correct results within the solver tolerances.

To evaluate SimBio's performance,
we selected \ac{SBML} models from the curated section of BioModels \cite{malik-sheriffBioModels15Years2020}.
Among the first 250 models,
we considered the 117 that used supported \ac{SBML} features.
We ran simulations
on a MacBook Air with M2 CPU
using Python 3.11.8,
COPASI v4.42.284
(with BasiCO v0.58) with the LSODA solver,
RoadRunner v2.5.0 (with Tellurium v2.2.10) with the comparable CVODE solver, and
SimBio v0.3.2 with the LSODA solver.
For SimBio, we considered three variants:
NumPy (v1.26.4) backend and scipy (v1.12.0)'s LSODA solver,
Numba (v0.59.0) backend and scipy's LSODA solver,
Numba backend and numbalsoda's (v0.3.5) LSODA solver.
In all cases, we used a relative tolerance of \(10^{-6}\) and absolute tolerances of \(10^{-9}\).
We measured two simulation stages:
an initial cold run that includes the reading of the \ac{SBML} model
and subsequent warm runs.

@@

The inclusion of \ac{SBML} support facilitates the effortless reuse of models created by the systems biology community,
along with the vast collection of public models hosted in the BioModels repository.
The modular architecture of these packages facilitates their reuse, enhancement, and extension by the wider Python community.
Therefore, it should be also easy to integrate with existing infrastructure such a BioSimulators.org  \cite{shaikhBioSimulatorsCentralRegistry2022},
or combine with other packages like SimService \cite{sego_simservice_2024} to build more complex simulations.
For instance, an individual from outside the systems biology field could contribute a stochastic integrator to poincaré,
which would then be available in SimBio.
This clear separation of concerns also makes the packages more comprehensible,
lowering the barrier for contributing improvements or new features.
Such an architecture ensures their maintainability and ongoing development well into the future.

@@

  \begin{minipage}[b]{0.50\linewidth}
  {
      \centering 
      \includegraphics[width=\textwidth]{src/ide/ide1.pdf}
  }
  \end{minipage}%
  %
  \begin{minipage}[b]{0.50\linewidth}
  {
      \centering
      \includegraphics[width=\textwidth]{src/ide/ide2.pdf}
  }
  \end{minipage}%

@@

  \begin{CodeOutput}
    \includegraphics[width=\textwidth]{article_files/cell-9-output-1}
  \end{CodeOutput}

@@

% FIGURE 6
\begin{figure*}[t]
  \centering
  \includegraphics[width=\textwidth]{src/performance/figures/performance}
  \caption{
    Performance of different softwares to solve models from the curated section of BioModels.
    (left) Run time for the model BIOMD3 as a function of the number of output points.
    (right) Run time for different models for 300 output points,
    using the geometric mean of the different softwares to order them.
    Each point corresponds to the median of 20 runs,
    with a neglibible errorbar given by the interquartile range.
  }
  \label{fig-runtime}
\end{figure*}

>>>>

Several tools already exist to \emph{define}, \emph{translate} and \emph{solve} \acp{CRN}.
BioSimulators.org \citep{shaikhBioSimulatorsCentralRegistry2022},
a registry of simulation tools,
lists at least 15 softwares categorized under Python including
COPASI \citep{hoopsCOPASICOmplexPAthway2006},
Tellurium \citep{choiTelluriumExtensiblePythonbased2018} and
PySB \citep{lopezProgrammingBiologicalModels2013}.
COPASI is a standalone software with a \ac{GUI}
that is widely used for its user-friendly interface and comprehensive features.
It also includes Python bindings, BASICO \citep{bergmannBASICOSimplifiedPython2023b}, that allow advanced scripting.
Tellurium is a Python-based modeling environment
that uses a C++ library called libRoadRunner \citep{welshLibRoadRunnerHighPerformance2023} in the backend to translate and solve models.
PySB is a Python library that created a \ac{DSL} using standard Python to define models,
which are then translated to \acp{ODE} using a Perl library called BioNetGen \citep{harrisBioNetGenAdvancesRulebased2016}.

One limitation of these tools is their extensibility from Python.
As they wrap libraries in other languages for defining, translating and/or solving models,
these steps cannot be altered or easily inspected from Python.
While they enable model definition and running simulations via Python scripts,
they cannot fully leverage Python's extensive package ecosystem.
For example,
COPASI and Tellurium do not allow the use of solvers defined in other Python packages,
and adding new integrators requires working with C++.
In particular, the step that translates into equations is not exposed by any of these tools.
As such, it is not possible to apply custom optimizations to the equations
or use automatic differentation packages such as JAX \citep{jax2018github} to compute the model's jacobian.

Another challenge is the way models are defined.
Many tools support the \ac{SBML} \citep{huckaSBMLL3V2} as an exchange format,
a \emph{de facto} standard for \acp{CRN} that defines species, parameters and reactions between species.
As writing \ac{SBML} directly is impractical,
Tellurium uses a \ac{DSL} called Antimony \citep{smithAntimonyModularModel2009} for defining models.
\Acp{DSL} allows to reuse the same code in different programming environments,
but are not recognized by default in \acp{IDE}
and, therefore, they cannot provide syntax highlighting, code completion, refactoring, and static analysis.
For Antimony, an extension providing these capabilities was developed for Visual Studio Code,
but its maintenance could be a demanding task for the systems biology community.
In the case of PySB,
using Python's dynamic nature,
it developers designed a \ac{DSL} within Python.
To save keystrokes,
it uses the global scope to create species and parameters,
without explicitly assigning them to Python variables or to the model,
but this approach is not fully compatible with \acp{IDE},
affecting the development experience.

@@

Modular code architecture makes code reusable, extensible, and easier to maintain.
Therefore, we split the code into three Python packages:
symbolite, to create symbolic expressions;
poincaré, to define dynamical systems;
and simbio, to define \acp{CRN} and interface with systems biology standards such as \ac{SBML}.
These are pure Python packages with standard dependencies from the PyData scientific stack
such as NumPy \citep{harrisArrayProgrammingNumPy2020} and pandas \citep{mckinneyDataStructuresStatistical2010}.
They are published in the \acf{PyPI},
where links to the source code and documentation hosted in GitHub can be found,
and can be easily installed with \texttt{pip\ install\ simbio},
which installs symbolite and poincare as dependencies.

Symbolite is a lightweight symbolics package to create algebraic mathematical expressions.
Unlike SymPy \citep{meurerSymPySymbolicComputing2017},
a widely-used Python library for symbolic mathematics,
it only provides the building of an expression tree
which can be inspected and compiled to various backends.
Symbolite is designed to facilitate the integration of new backends.
Currently, we have implementations for
NumPy \citep{harrisArrayProgrammingNumPy2020};
Numba \citep{lamNumbaLLVMbasedPython2015},
a \ac{JIT} compiler to LLVM;
SymPy \citep{meurerSymPySymbolicComputing2017};
and JAX \citep{jax2018github}, a
library that support automatic differentiation and compilation to \acp{GPU} and \acp{TPU}.

@@

For this last point,
\acp{IDE} that support \texttt{dataclass\_transform} \citep{debontePEP681Data2021}
can provide a tooltip with the expected signature (Figure~\ref{fig-ide}).
This requires the use of type annotations,
which play a more significant role in static type checking
as they can help to identify errors before running the code.
For instance, to parameterize the initial conditions of variables
we have to use a \texttt{Constant}.
If we try to use a \texttt{Parameter},
which could be a time-dependent expression,
it is flagged as a type error (Figure~\ref{fig-ide}).

@@

Several commonly used reactions are predefined as \texttt{MassAction} subclasses,
such as \texttt{MichaelisMenten} (\(S + E \leftrightarrow ES \rightarrow P + E\))
and its approximate form without the intermediate species \(ES\),
and it is also simple to implement used-defined ones
as subclasses of \texttt{RateLaw} or \texttt{MassAction}.
Additionally, SimBio supports importing models from \ac{SBML},
and downloading them directly from BioModels \citep{malik-sheriffBioModels15Years2020} (Figure~\ref{fig-simbio-io}).
Work is in progress to support exporting to \ac{SBML} and add more \ac{SBML} features.
Currently \ac{SBML} unsupported features include
algebraic rules, constraints, events, reactions with the fast attribute or with math stoichiometry,
units and compartments with size different from 1.

@@

To evaluate SimBio's reproducibility,
we used the \ac{SBML} test suite \citep{SBMLTestSuite},
which provides a set of \ac{SBML} models
and the expected result of a simulation.
Excluding models that use \ac{SBML} features not yet supported by SimBio,
every simulation returned correct results within the solver tolerances.

To evaluate SimBio's performance,
we selected \ac{SBML} models from the curated section of BioModels \citep{malik-sheriffBioModels15Years2020}.
Among the first 250 models,
we considered the 117 that used supported \ac{SBML} features.
We ran simulations
on a MacBook Air with M2 CPU
using Python 3.11.8,
COPASI v4.42.284
(with BasiCO v0.58) with the LSODA solver,
RoadRunner v2.5.0 (with Tellurium v2.2.10) with the comparable CVODE solver, and
SimBio v0.3.2 with the LSODA solver.
For SimBio, we considered three variants:
NumPy (v1.26.4) backend and scipy (v1.12.0)'s LSODA solver,
Numba (v0.59.0) backend and scipy's LSODA solver,
Numba backend and numbalsoda's (v0.3.5) LSODA solver.
In all cases, we used a relative tolerance of \(10^{-6}\) and absolute tolerances of \(10^{-9}\).
We measured two simulation stages:
an initial cold run that includes the reading of the \ac{SBML} model
and subsequent warm runs.

@@

The inclusion of \ac{SBML} support facilitates the effortless reuse of models created by the systems biology community,
along with the vast collection of public models hosted in the BioModels repository.
The modular architecture of these packages facilitates their reuse, enhancement, and extension by the wider Python community.
Therefore, it should be also easy to integrate with existing infrastructure such a BioSimulators.org \citep{shaikhBioSimulatorsCentralRegistry2022},
or combine with other packages like SimService \citep{sego_simservice_2024} to build more complex simulations.
For instance, an individual from outside the systems biology field could contribute a stochastic integrator to poincaré,
which would then be available in SimBio.
This clear separation of concerns also makes the packages more comprehensible,
lowering the barrier for contributing improvements or new features.
Such an architecture ensures their maintainability and ongoing development well into the future.

@@

  \begin{minipage}[b]{0.50\linewidth}
  {
      \centering 
      \includegraphics[width=\textwidth]{ide1.pdf}
  }
  \end{minipage}%
  %
  \begin{minipage}[b]{0.50\linewidth}
  {
      \centering
      \includegraphics[width=\textwidth]{ide2.pdf}
  }
  \end{minipage}%

@@

  \begin{CodeOutput}
    \includegraphics[width=\textwidth]{cell-output}
  \end{CodeOutput}

@@

% FIGURE 6
\begin{figure*}[t]
  \centering
  \includegraphics[width=\textwidth]{performance}
  \caption{
    Performance of different softwares to solve models from the curated section of BioModels.
    (left) Run time for the model BIOMD3 as a function of the number of output points.
    (right) Run time for different models for 300 output points,
    using the geometric mean of the different softwares to order them.
    Each point corresponds to the median of 20 runs,
    with a neglibible errorbar given by the interquartile range.
  }
  \label{fig-runtime}
\end{figure*}
